A Constraint Solving Approach to Parikh Images of Regular Languages

### 2024-03-08 evaluation.tex
Add artefact, update evaluation

 We also attempted to benchmark instances generated by
\OstrichPlus{} solving the Kaluza benchmarks~\cite{Saxena10:kaluza}
(\numprint{38227}~instances),
\iffalse
and pyex-len
(\numprint{791}~instances) benchmark suites\fi
but discarded them since they all turned out to be trivial for our tool
\Catra{}: every instance
could be solved in under five seconds, with a mean runtime of about 0.1s.
The benchmarks are run
on commit~\texttt{\commit} of \Catra{}\footnote{Github repository redacted for anonymity.}.

The benchmarks are executed in parallel across a cluster of identical
machines, \BenchmarkRig{}, as the sole job for each machine.  We compiled
the code using Scala~\ScalaVersion{} and executed the experiments
on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{}
version~\NuxmvVersion{} invoked as a subprocess for each instance. Instances
were executed in batches of \BatchSize{}, each given a fresh JVM. Each JVM was
warmed up for \SI{10}{s} on a random benchmark from the set before starting
to execute. We believe this represents a realistic use case where \Calculus{} is
used to support, e.g., a string solver. Experiments were executed in random order
for all backends. Each instance got a time budget of~\RuntimeTimeout{}.

@@

\subsubsection{Scalability}\label{sec:scaling}%
To investigate scaling we additionally execute \numprint{2000} randomly sampled
(without replacement) benchmarks with a \numprint{120}-second timeout. We add
runs of \Calculus{} with clause learning and restarts disabled, and with only
clause learning disabled to show the impact of \cref{sec:random-restarts}, and
\cref{sec:clause-learning} respectively.

@@

The most obvious threat to validity would be an unsound implementation. To
address this we have validated all reported solutions made by \Calculus{} with
\Nuxmv{}. A previous version contained a race with random restarts during
product materialisation causing non-deterministic unsoundness in \numprint{0.7}
of instances. Additionally, both we and the artefact evaluation committee independently
discovered a soundness issue in CA-Str on one instance in QF\_S where the 
\texttt{QF\_S\slash{}20230329-automatark-lu\slash{}instance08425.smt2} instance was
incorrectly reported as \textsc{Unsat}. The underlying issue was that \Catra{} generated
a too broad blocking clause when encountering unsatisifiability stemming from 
an intermittent product accepting only the empty string during product materialisation. 
We implemented a fix for that bug and re-evaluated all benchmarks in SMT-COMP.

Since addressing those bugs we have observed no further soundness
issues in neither \Catra{} nor CA-Str. Benchmarking results for both runs are included
in the artefact of this paper, and show virtually identical performance characteristics.

The second threat to validity is our implementation of automata operations. As
\Calculus{} by design offloads some of the product computation work onto
\Princess{}, the baseline could be unfairly disadvantaged by a slow automata
product implementation. We believe this is not an issue since similar
performance issues with the baseline approach have been reported for other string
solvers, as well as for a previous implementation in \Ostrich{}. Additionally,
profiling shows that baseline spends most of its time in
\Princess{}, suggesting that the automaton implementation is not the bottleneck.
Finally, the difference in performance between \Calculus{} and baseline has been
robust under significant optimisation of the automata library, additionally
strengthening this thesis.

>>>>

 We also attempted to benchmark instances generated by
\OstrichPlus{} solving the Kaluza benchmarks~\cite{Saxena10:kaluza}
(\numprint{38227}~instances),
\iffalse
and pyex-len
(\numprint{791}~instances) benchmark suites\fi
but discarded them since they all turned out to be trivial for our tool
\Catra{}: every instance
could be solved in under five seconds, with a mean runtime of about 0.1s.
The benchmarks are run
on commit~\texttt{\commit} of \Catra{}.

The benchmarks for \Calculus{} and \Nuxmv{} are executed in parallel on \BenchmarkRig{}, with 4~threads
sharing the same JVM. Baseline was
executed one JVM instance per input as 6~jobs in with a \SI{4}{GB} heap each on an Intel i5-10600, \SI{3.3}{GHz} CPU
with 6 cores and 12 threads. Simultaneously, StarExec ran the same experiment and produced almost identical results.

We compiled
the code using Scala~\ScalaVersion{} and executed the experiments
on~\JvmVersion{} with a maximum heap of~\MaxHeapSize{}. We used \Nuxmv{}
version~\NuxmvVersion{} invoked as a subprocess for each instance. Instances
were executed in batches of \BatchSize{}, each given a fresh JVM. Each JVM was
warmed up for \SI{10}{s} on a random benchmark from the set before starting
to execute. We believe this represents a realistic use case where \Calculus{} is
used to support, e.g., a string solver. Experiments were executed in random order
for all backends. Each instance got a time budget of~\RuntimeTimeout{}.

@@

\subsubsection{Scalability}\label{sec:scaling}%
To investigate scaling we additionally execute \numprint{2000} randomly sampled
(without replacement) benchmarks with a \numprint{120}-second timeout. We add
runs of \Calculus{} with clause learning and restarts disabled, and with only
clause learning disabled to show the impact of \cref{sec:random-restarts}, and
\cref{sec:clause-learning} respectively. These experiments are executed on a smaller
machine, a six-core AMD~Ryzen~5 2600 but the same Java version. Their max
heap size is set to \SI{20}{GB} since this system has less RAM. The baseline
experiments are executed separately as only two threads.

@@

The most obvious threat to validity would be an unsound implementation. To
address this we have validated all reported solutions made by \Calculus{} with
\Nuxmv{}. A previous version contained a race with random restarts during
product materialisation causing non-deterministic unsoundness in \numprint{0.7}
of instances. Additionally, both we and the artefact evaluation committee independently
discovered a soundness issue in CA-Str on one instance in QF\_S where the 
\texttt{20230329-automatark-lu\slash{}instance08425.smt2} instance was
incorrectly reported as \textsc{Unsat}. The underlying issue was that \Catra{} generated
a too broad blocking clause for a product without transitions. We implemented a fix
and re-evaluated all benchmarks. Performance was not measurably affected.

Since addressing those bugs we have observed no further soundness
issues in neither \Catra{} nor in CA-Str. Benchmarking results for both runs are included
in the artefact of this paper, and show virtually identical performance characteristics. We have
additionally executed all of the benchmarks on machines with widely different performance
characteristics and have observed the trends to be robust.

The second threat to validity is our implementation of automata operations. As
\Calculus{} by design offloads some of the product computation work onto
\Princess{}, the baseline could be unfairly disadvantaged by a slow automata
product implementation. We believe this is not an issue since similar
performance issues with the baseline approach have been reported for other string
solvers, as well as for a previous implementation in \Ostrich{}. Additionally,
profiling shows that baseline spends most of its time in
\Princess{}, suggesting that the automaton implementation is not the bottleneck.
Finally, the difference in performance between \Calculus{} and baseline was unaffected
 by significant optimisation of the automata library, additionally
strengthening this thesis.

### 2024-03-08 implementation.tex
Add artefact, update evaluation

\Catra{} is written in Scala, with the calculus described in this paper
implemented as a theory plug-in for the \Princess{} automated theorem
prover~\cite{princess}, which also performs the Presburger reasoning. For
comparison, we also provide an implementation of the baseline method
from~\cite{generate-parikh-image}, a direct translation that uses the~\Nuxmv{}
symbolic model checker~\cite{nuxmv} to solve our constraints, and the
approximation described in~\cite{approximate-parikh} on top of the standard
baseline back-end. An example of an input file corresponding to our running
example introduced in \cref{sec:intuition} can be found in the root directory
of the artefact~\cref{artefact}.

>>>>

\Catra{} is written in Scala, with the calculus described in this paper
implemented as a theory plug-in for the \Princess{} automated theorem
prover~\cite{princess}, which also performs the Presburger reasoning. For
comparison, we also provide an implementation of the baseline method
from~\cite{generate-parikh-image}, a direct translation that uses the~\Nuxmv{}
symbolic model checker~\cite{nuxmv} to solve our constraints, and the
approximation described in~\cite{approximate-parikh} on top of the standard
baseline back-end. An example of an input file corresponding to our running
example introduced in \cref{sec:intuition} can be found in the root directory
of the artefact~\cite{artefact}.
